Typed Functional Genetic Programming

### 2013-07-20 tex/thesis.tex
[tex]

\theoremstyle{plain} 
\newtheorem{theorem}{Theorem} 
\newtheorem{proposition}{Proposition} 
\newtheorem{lemma}{Lemma} 
\newtheorem*{corollary}{Corollary}

@@

\begin{todo}
& zmínit $=_\beta$ a $\twoheadrightarrow_\beta$ 
jestli se k tomu někde dál v textu odkazuju
\end{todo}

@@

@@

\newcommand{\ruleI}[2]{\dfrac{#1}{#2}}
\newcommand{\ruleII}[3]{\dfrac{#1 \qquad #2}{#3}}
\newcommand{\ruleIII}[4]{\dfrac{#1 \qquad #2 \qquad #3}{#4}}
\newcommand{\ruleIV}[5]{\dfrac{#1 \qquad #2 \qquad #3 \qquad #4}{#5}}

@@

\section{Term generating grammar}
Inference rules are good for deriving statements of the form \GMS, but our
goal is slightly different; we would like to generate many \lterms M for a given type 
$\sigma$ and context $\Gamma$.\\

@@

%It corresponds to following grammar rule schema.
%$$ ( \sigma , \Gamma )  \longmapsto \mbox{ {\Large x}} 
%~~~~~~~~~~~~~ \textbf{if } (x,\sigma) \in \Gamma$$

@@

For every non-terminal $(\tau,\Gamma)$ 
and for every type $\sigma \in \T$ 
there is following grammar rule.\footnote{ 
Terminal symbols for parenthesis and normally {\it space} 
now \textvisiblespace \quad (for {\it function application} operator) 
are visually highlighted. } 
$$
	( \tau , \Gamma )  \longmapsto
	\bigg( ( \sigma \rightarrow \tau , \Gamma ) 
	\mbox{ \Vtextvisiblespace[1em] } ( \sigma , \Gamma ) \bigg)
$$\\

@@

$$ 
	( \sigma \rightarrow \tau , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large $\lambda$ x . }}( \tau , \Gamma \cup \{ (x,\sigma) \} ) \quad \bigg)
$$\\

We will demonstrate \lterm generation on example. 
Again on $(\lambda f . (\lambda x . (f x) ))$. 
We would like to generate \lterm of a type 
$(\sigma \rightarrow \tau) \rightarrow (\sigma \rightarrow \tau)$
with $\Gamma = \{\}$.
\begin{align*}
	& ((\sigma \rightarrow \tau) \rightarrow (\sigma \rightarrow \tau),\{\}) \\ 
	\longmapsto & \Big( \mbox{ {\Large $\lambda$f.}}
	  ( \sigma \rightarrow \tau , \{ (f,\sigma \rightarrow \tau) \} ) 
	~ \Big)
	\\
	\longmapsto & 
	\Big( \mbox{ {\Large $\lambda$f. }}
		\Big( \mbox{ {\Large $\lambda$x. }}
	  	 	( \tau , \{ (f,\sigma \rightarrow \tau) , (x,\sigma) \} ) 
		~ \Big)  	 
	~ \Big)
	\\
	\longmapsto & 
	\Big( \mbox{ {\Large $\lambda$f. }}
		\Big( \mbox{ {\Large $\lambda$x. }}	  	 	
	  	 	\Big( 
	  	 	  ( \sigma \rightarrow \tau , \{ (f,\sigma \rightarrow \tau) , (x,\sigma) \} ) 
			  \mbox{ \Vtextvisiblespace[1em] } 
			  ( \sigma , \{ (f,\sigma \rightarrow \tau) , (x,\sigma) \} )  \Big) 
		~ \Big)  	 
	 ~ \Big)
	\\
	\longmapsto & 
	\Big( \mbox{ {\Large $\lambda$f. }}
		\Big( \mbox{ {\Large $\lambda$x. }}	  	 	
	  	 	\Big( 
	  	 	  \mbox{ {\Large f}} 
			  \mbox{ \Vtextvisiblespace[1em] } 
			  ( \sigma , \{ (f,\sigma \rightarrow \tau) , (x,\sigma) \} ) \Big) 
		~ \Big)  	 
	~ \Big)		
	\\
	\longmapsto & 
	\Big( \mbox{ {\Large $\lambda$f. }}
		\Big( \mbox{ {\Large $\lambda$x. }}	  	 	
	  	 	\Big( 
	  	 	  \mbox{ {\Large f}} 
			  \mbox{ \Vtextvisiblespace[1em] } 
			  \mbox{{\Large x}} \Big) 
		~ \Big)  	 
	~ \Big)
\end{align*}~\\

@@

The corresponding grammar rule can be summarized as follows.

$$
 (\sigma_{n+1},\Gamma_{n+1})
 \longmapsto
 \mbox{ {\Large \textbf{T}}} 
 \bigg( 
 (\sigma_1,\Gamma_1)
 \mbox{ {\Large ,}}
 (\sigma_2,\Gamma_2)
 \mbox{ {\Large ,}}
 ~\dots
 \mbox{ {\Large ,}}
 (\sigma_n,\Gamma_n)
 \bigg)  
$$

\begin{todo}
   & asi ocitovat toho van Wijngaarden [1981] s těma 2-level grammar
     jak je to v Barendregtovi \cite{barendregt10} 
   & vyřešit jak je to stou freš proměnnou x v introdukci...
     && stejně tak v tom obecnym případě to dělá neplechu
\end{todo}
%	( \tau , \Gamma )  \longmapsto
%	( \sigma \rightarrow \tau , \Gamma )  \longmapsto
%	( \sigma , \Gamma )  \longmapsto \mbox{ {\Large x}}
%	\longmapsto & \Big( \mbox{ {\Large $\lambda$f.}}
%	\longmapsto & 
%	\longmapsto & 
%	\longmapsto & 
%	\longmapsto & 

@@

Let \GMS where 
$\sigma = \tau_0 \ar \tau_1 \ar \dots \ar \tau_n \ar \alpha$,
$n \geq 0$.\\
Then $M$ is in \textit{long normal form} if following 
conditions are satisfied:

\begin{itemize}
 \item $M$ is term of a form $\lam{x_1 \dots x_n}{f~M_1~\dots~M_n}$,
 \item each $M_i$ is in \textit{long normal form}.
\end{itemize}

@@

\begin{todo}
 & popsat tu jak ji předelat na beta nf
	pokud by se nám chtělo... 
	V barendrech84 (ta nová kniha s žlutejma deskama) je ukázano že je 
	eta CR (čurč roserovská) což snad stačí společně s 
	tim tvrzeničkem z book.pdf že vezmu lnf
	a dělám eta redukce dokud to de. až to nejde tak bych to měl mít v 
	beta-nf
 & přepsat to nějak hezčejc, ten seznam heskejch vlastností
\end{todo}

%"Barendregt-like"
\section{Inference and grammar rules producing terms in Long normal form}
\label{barlike}

As is said in \cite{barendregt10} there are two
inference/grammar rules which together generate
simply typed lambda terms in their long normal forms.

Inference rule 1: 
\[
	\frac{\Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} \vdash M : \alpha }
	     {\Gamma \vdash (\lambda x_1 \dots x_n . M) : 
	     \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha }
\]
\\
Proof of correctness:
\[
	\dfrac{
		\dfrac
		 {\Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} \vdash M : \alpha}
		 {\dfrac
		   {\Gamma \cup \{ (x_1,\tau_1),\dots,(x_{n-1},\tau_{n-1})\} \cup 
		                \{(x_n,\tau_n) \} \vdash M : \alpha}
		   {\dfrac{\Gamma \cup \{ (x_1,\tau_1),\dots,(x_{n-1},\tau_{n-1})\}  
		                \vdash (\lambda x_n . M) : \tau_n \rightarrow \alpha}
				  { \vdots }		   
		   }
		 }		 
	 }
	     {\Gamma \vdash (\lambda x_1 \dots x_n . M) : 
	     \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha }
\]
\\
... there is a grammar rule:
\[ 
	( \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large 
	$\lambda x_1 \dots x_n .$ 
	}}( \alpha , \Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} ) \quad \bigg)
\]
\\

Inference rule 2: 
\[
	\frac{ (f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma \qquad
	       \Gamma \vdash M_1 : \rho_1 \quad
	       \dotsm \quad
	       \Gamma \vdash M_m : \rho_m        
	      }
	     {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
Proof of correctness (\textbf{TODO REPAIR} Conceptually it is ok but there is sazba-bug somewhere): 
\[
   \dfrac
     {\dfrac
      {\dfrac
       {\dfrac         
         {\dfrac  
          {\dfrac
           {\boxed{(f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma}}
           {\Gamma \vdash f : \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha}
           \quad
           \boxed{\Gamma \vdash M_1 : \rho_1} }
          {\Gamma \vdash (f M_1) : \rho_2 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha }
          }{\vdots} 
         \quad 
         \ddots }
       {\Gamma \vdash (f M_1 \dots M_{m-2}) : \rho_{m-1} \rightarrow \rho_m \rightarrow \alpha}
       \quad
       \boxed{\Gamma \vdash M_{m-1} : \rho_{m-1}}  }
      {\Gamma \vdash (f M_1 \dots M_{m-1}) : \rho_m \rightarrow \alpha}       
      \quad 
      \boxed{\Gamma \vdash M_m : \rho_m} }
	 {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
... there is a grammar rule:
\[ 
	( \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]

\begin{todo}
& SHOW correctness of those inference rules by composing them of 
	  $E^{\rightarrow}$, $I^{\rightarrow}$ and \textit{axiom}.
& SHOW more examples of inference rules transformed into grammar rules.
& DESCRIBE general algorithm for this transformation.
& TALK ABOUT $\tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$ 
& TALK ABOUT $\beta \eta^{-1}$-normal form which is generated by this method.
& Napsat tu (nebo asi někam jinam
	ale hlavně někam), že de generovat přímo i termy v beta normalní 
	formě, ale že se mi to zdálo o dost pracnější díky tomu, že
	musíme čekovat jestli neni něco toho typu i v ne atomickejch
	vrcholech/typech...
\end{todo}

@@

Compare this case with corresponding grammar rule:
\[ 
	( \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]

@@

Compare this case with corresponding grammar rule:
\[ 
	( \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large 
	$\lambda x_1 \dots x_n .$ 
	}}( \alpha , \Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} ) \quad \bigg)
\]
~\\
\Lets consider following $(\sigma,\Gamma)$ as a simple example:

@@

\chapter{Ideas yet not implemented}

@@

>>>>

\theoremstyle{plain} 
\newtheorem{theorem}{Theorem} 
\newtheorem*{proposition}{Proposition} 
\newtheorem{lemma}{Lemma} 
\newtheorem*{corollary}{Corollary}

@@

\begin{todo}
& !!!!!!! zmínit $=_\beta$ a $\twoheadrightarrow_\beta$  a i pro ostatní redukce!!!!!
\end{todo}

@@

\begin{todo}
 & zapsat sem notaci s $\Gamma,x:\sigma $ denotes $ \Gamma\cup\{(x:\sigma)\}$ 
\end{todo}

@@

\newcommand{\ruleI}[2]{\dfrac{#1}{#2}}
\newcommand{\ruleII}[3]{\dfrac{#1 \qquad #2}{#3}}
\newcommand{\ruleIIs}[3]{\dfrac{#1 ~ #2}{#3}}
\newcommand{\ruleIII}[4]{\dfrac{#1 \qquad #2 \qquad #3}{#4}}
\newcommand{\ruleIV}[5]{\dfrac{#1 \qquad #2 \qquad #3 \qquad #4}{#5}}

@@

\section{Term generating grammar}

\newcommand{\gar}{\longmapsto}

Inference rules are good for deriving statements of the form \GMS, but our
goal is slightly different; we would like to generate many \lterms M for a given type 
$\sigma$ and context $\Gamma$.\\

@@

%It corresponds to following grammar rule schema.
%$$ ( \sigma , \Gamma )  \gar \mbox{ {\Large x}} 
%~~~~~~~~~~~~~ \textbf{if } (x,\sigma) \in \Gamma$$

@@

For every non-terminal $(\tau,\Gamma)$ 
and for every type $\sigma \in \T$ 
there is following grammar rule.\footnote{ 
Terminal symbols for parenthesis and normally {\it space} 
now \textvisiblespace \quad (for {\it function application} operator) 
are visually highlighted. } 
$$
	( \tau , \Gamma )  \gar
	\bigg( ( \sigma \rightarrow \tau , \Gamma ) 
	\mbox{ \Vtextvisiblespace[1em] } ( \sigma , \Gamma ) \bigg)
$$\\

@@

$$ 
	( \sigma \rightarrow \tau , \Gamma )  \gar
	\bigg( \mbox{ {\Large $\lambda$ x . }}( \tau , \Gamma \cup \{ (x,\sigma) \} ) \quad \bigg)
$$\\

We will demonstrate \lterm generation on example. 
Again on $(\lambda f . (\lambda x . (f x) ))$. 
We would like to generate \lterm of a type 
$(\sigma \rightarrow \tau) \rightarrow (\sigma \rightarrow \tau)$
with $\Gamma = \{\}$.
\begin{align*}
	& ((\sigma \rightarrow \tau) \rightarrow (\sigma \rightarrow \tau),\{\}) \\ 
	\gar & \Big( \mbox{ {\Large $\lambda$f.}}
	  ( \sigma \rightarrow \tau , \{ (f,\sigma \rightarrow \tau) \} ) 
	~ \Big)
	\\
	\gar & 
	\Big( \mbox{ {\Large $\lambda$f. }}
		\Big( \mbox{ {\Large $\lambda$x. }}
	  	 	( \tau , \{ (f,\sigma \rightarrow \tau) , (x,\sigma) \} ) 
		~ \Big)  	 
	~ \Big)
	\\
	\gar & 
	\Big( \mbox{ {\Large $\lambda$f. }}
		\Big( \mbox{ {\Large $\lambda$x. }}	  	 	
	  	 	\Big( 
	  	 	  ( \sigma \rightarrow \tau , \{ (f,\sigma \rightarrow \tau) , (x,\sigma) \} ) 
			  \mbox{ \Vtextvisiblespace[1em] } 
			  ( \sigma , \{ (f,\sigma \rightarrow \tau) , (x,\sigma) \} )  \Big) 
		~ \Big)  	 
	 ~ \Big)
	\\
	\gar & 
	\Big( \mbox{ {\Large $\lambda$f. }}
		\Big( \mbox{ {\Large $\lambda$x. }}	  	 	
	  	 	\Big( 
	  	 	  \mbox{ {\Large f}} 
			  \mbox{ \Vtextvisiblespace[1em] } 
			  ( \sigma , \{ (f,\sigma \rightarrow \tau) , (x,\sigma) \} ) \Big) 
		~ \Big)  	 
	~ \Big)		
	\\
	\gar & 
	\Big( \mbox{ {\Large $\lambda$f. }}
		\Big( \mbox{ {\Large $\lambda$x. }}	  	 	
	  	 	\Big( 
	  	 	  \mbox{ {\Large f}} 
			  \mbox{ \Vtextvisiblespace[1em] } 
			  \mbox{{\Large x}} \Big) 
		~ \Big)  	 
	~ \Big)
\end{align*}~\\

@@

The corresponding grammar rule can be summarized as following schema.

$$
 (\sigma_{n+1},\Gamma_{n+1})
 \gar
 \mbox{ {\Large \textbf{T}}} 
 \bigg( 
 (\sigma_1,\Gamma_1)
 \mbox{ {\Large ,}}
 (\sigma_2,\Gamma_2)
 \mbox{ {\Large ,}}
 ~\dots
 \mbox{ {\Large ,}}
 (\sigma_n,\Gamma_n)
 \bigg)  
$$

\begin{todo}
   & buď dokázat ty převody na grammar rules explicitně
     nebo říct že se to dokáže podobnym stylem jako v dukazu 
     v sekci \ref{barlike}
   & možná ocitovat toho van Wijngaarden [1981] s těma 2-level grammar
     jak je to v Barendregtovi \cite{barendregt10} 
   & vyřešit jak je to stou freš proměnnou x v introdukci...
     && stejně tak v tom obecnym případě to dělá neplechu
\end{todo}

@@

\newcommand{\lnf}{\textit{lnf}\xspace}

\begin{definition}
Let \GMS where 
$\sigma = \tau_1 \ar \dots \ar \tau_n \ar \alpha$,
$n \geq 0$.
Then $M$ is in \textit{long normal form} (\textit{lnf}) if following 
conditions are satisfied.
\begin{enumerate}
 \item $M$ is term of the form $\lam{x_1 \dots x_n}{f~M_1~\dots~M_m}$\\
  (specially for $n = 0$, $M$ is term of the form $f$),
 \item each $M_i$ is in \textit{long normal form}.
\end{enumerate}
\end{definition}~

@@

\begin{todo}
 & v barendrechtovi je to v definici lnf $\lam{x_1 \dots x_n}{f~M_1~\dots~M_n}$ místo 
   $M_m$ ale to musí bejt určitě překlep...
 & popsat tu jak ji předelat na beta nf
	pokud by se nám chtělo... 
	V barendrech84 (ta nová kniha s žlutejma deskama) je ukázano že je 
	eta CR (čurč roserovská) což snad stačí společně s 
	tim tvrzeničkem z book.pdf že vezmu lnf
	a dělám eta redukce dokud to de. až to nejde tak bych to měl mít v 
	beta-nf
 & přepsat to nějak hezčejc, ten seznam heskejch vlastností
\end{todo}

\section{Grammar producing \lterms in \lnf}
\label{barlike}

In \cite{barendregt10} is shown term generating grammar with 
following rules (our notation is used, but we will not 
highlighted terminals anymore).\footnote{
I was using term generating grammars for term generation
on my own before I encountered with \cite{barendregt10},
where i happily discovered that Barendregt is using
almost identical notation.
}
\begin{align*}
( \alpha , \Gamma )  
&\gar
(~f~( \rho_1 , \Gamma )~\dots~( \rho_m , \Gamma )~)
& \textbf{if } \alpha \in A,
(f : \rho_1 \ar \dots \ar \rho_m \ar \alpha) \in \Gamma
\\ 
( \sigma \rightarrow \tau , \Gamma )  
&\gar
(~\lambda~x~.~( \tau ; \Gamma,x:\sigma )~)
&   
\end{align*}

The second rule can be replaced by more effective one.
\[ 
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
\gar
(~\lambda~x_1~\dots~x_n~.~
( \alpha ; \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n  )~)
\]

This rule packs consecutive uses of the second rule into one use.
This is valid since the use of the second rule is deterministic;
it is used if and only if the non-terminal's type is not atomic.\\

Following proposition about this grammar is stated (but not proven)
in~\cite{barendregt10}. Since it is key proposition for our 
term generating method, we also provide a proof.

\newcommand{\garr}{\leadsto}%{\Rrightarrow}%
\newcommand{\Gp}{\Gamma^\prime}

\begin{proposition}
Let context $\Gamma$, \lterm $M$ and type $\sigma$ be given.\\
Let $\garr$ be transitive reflexive closure of $\gar$. 
Then
$$ 
(\sigma,\Gamma) \garr M 
\Leftrightarrow
\tur{\Gamma}{M}{\sigma} \text{ and $M$ is in lnf}.   
$$
\end{proposition}

\begin{proof}~

\Lets start with proving

$ (\sigma,\Gamma) \garr M 
\then
\tur{\Gamma}{M}{\sigma} \text{ and $M$ is in \textit{lnf}}$

by induction on the length of rewriting chain of $(\sigma,\Gamma) \garr M$.\\

Let $\sigma = \tau_1 \ar \dots \ar \tau_n \ar \alpha (n \geq 0)$,

$\Gp = \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n$.\\

$(\tau_1 \ar \dots \ar \tau_n \ar \alpha,\Gamma), n>0$ must rewrite to
$(~\lambda~x_1~\dots~x_n~.~( \alpha , \Gp )~)$,

thus (even for $n = 0$) the chain must go through
$(~\lambda~x_1~\dots~x_n~.~( \alpha , \Gp )~)$.\\

$(\alpha,\Gp)$ must rewrite to 
$(~f~( \rho_1 , \Gp )~\dots~( \rho_m , \Gp )~)$

for some $(f,\rho_1\ar\dots\ar\rho_m\ar\alpha) \in \Gp$,

thus the chain must go through 
$(~\lambda~x_1~\dots~x_n~.~(~f~( \rho_1 , \Gp )~\dots~( \rho_m , \Gp )~)~)$.\\

After that each $( \rho_i , \Gp ) \garr M_i$
with shorter chain than $(\sigma,\Gamma) \garr M $.

Thus we have from the induction hypothesis for each $M_i$ that

$\tur{\Gp}{M_i}{\rho_i}$ and $M_i$ is in \textit{lnf}.\\

$M = (~\lambda~x_1~\dots~x_n~.~(~f~M_1~\dots~M_m~)~)$ 
and each $M_i$ is in \lnf,

thus M is in \lnf.\\

Following derivation shows how to derive \GMS.\\

$
\ruleI{
\ruleI{ 
\ruleI{
\ruleI{
\ruleIIs{
\ruleIIs{
\ruleIIs{
\ruleIIs{ 
\ruleI
{\boxed{(f,\rho_1\ar\dots\ar\rho_m\ar\alpha) \in \Gp} }
{\tur{\Gp}{f}{\rho_1\ar\dots\ar\rho_m\ar\alpha}}}
{\boxed{
 \tur{\Gp }
     {M_1 }
     {\rho_1 }}}
{
\tur{\Gp}{(~f~M_1~)}{\rho_2\ar\dots\ar\rho_m\ar\alpha} 
}}
{\boxed{
 \tur{\Gp }
     {M_2 }
     {\rho_2 }}}
{\vdots}}
{\ddots}
{\tur{\Gp }
     {(~f~M_1~\dots~M_{m-1}~) }
     {\rho_m \ar \alpha}}}
{\boxed{
 \tur{\Gp }
     {M_m }
     {\rho_m}} }
{\tur{\Gp }
     {(~f~M_1~\dots~M_m~) }
     {\alpha}}}
{\tur{\Gamma,x_1:\tau_1,\dots,x_{n-1}:\tau_{n-1} }
     {\lam{x_n}{(~f~M_1~\dots~M_m~)} }
     {\tau_n \ar \alpha} }}
{\vdots}}
{\tur{\Gamma,x_1:\tau_1 }
       {\lam{x_2~\dots~x_n}{(~f~M_1~\dots~M_m~)} }
       {\tau_2 \ar \dots \ar \tau_n \ar \alpha} } }
{\boxed{
 \tur{\Gamma }
     {\lam{x_1~x_2~\dots~x_n}{(~f~M_1~\dots~M_m~)} }
     {\tau_1 \ar \tau_2 \ar \dots \ar \tau_n \ar \alpha} }}
$

~\\~\\

Now \lets prove the opposite direction

$\tur{\Gamma}{M}{\sigma} \text{ and $M$ is in \textit{lnf}}
\then
(\sigma,\Gamma) \garr M$

by induction on the size of M.\\

Let $M = \lam{x_1 \dots x_n}{f~M_1~\dots~M_m}$,

$\sigma = \tau_1 \ar \dots \ar \tau_n \ar \alpha$,

$\Gp = \Gamma , x_1:\tau_1 , \dots , x_n:\tau_n$.\\

Then

$
( \sigma , \Gamma ) =
( \tau_1 \ar \dots \ar \tau_n \ar \alpha , \Gamma )  
\gar
(~\lambda~x_1~\dots~x_n~.~
( \alpha , \Gp )~)
$

by using the updated second grammar rule.\\

From \GMS we have that $(f,\rho_1\ar\dots\ar\rho_m\ar\alpha) \in \Gp$
and $\tur{\Gp}{M_i}{\rho_i}$.

(This can be easily observed from definition of \GMS.)\\

Therefore

$
(~\lambda~x_1~\dots~x_n~.~( \alpha , \Gp )~)
\gar
(~\lambda~x_1~\dots~x_n~.~
	(~f~( \rho_1 , \Gp )~\dots~( \rho_m , \Gp )~)
~)
$

by using the first rule.\\

Each $M_i$ is in \lnf, 
is smaller in size then $M$
and $\tur{\Gp}{M_i}{\rho_i}$;

therefore we have from the induction hypothesis that 
$(\rho_i,\Gp) \garr M_i$.\\

Therefore 
$
( \sigma , \Gamma )
\garr
(~\lambda~x_1~\dots~x_n~.~
	(~f~M_1~\dots~M_m~)
~)
$.

\end{proof}~

This means that we can use this grammar to generate all \lterms $M$ in \lnf
with desired type $\sigma$ from \textit{building blocks} context $\Gamma$.\\

\subsection{Benefits of generating \lterms in \lnf}

By generating \lterms in \textit{lnf} we avoid generating 
\lterms $M$,$N$ such that $M \not= N$, but $M =_{\beta\eta} N$.
In other words, we avoid generating two programs with different 
source codes, but performing the same computation.

Every \lterm $M$ such that \GMS has its unique \lnf $L$, 
for which $L =_{\beta\eta} M$.
Therefore the computation performed by \lterm $M$ 
is not omitted, because it is the same computation
as the computation performed by \lterm $L$. \\

\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}

Generating \lterms in \lnf is even better than generating 
\lterms in \bnf. Since \lnf is the same thing as \beenf,
every \lterm in \lnf is also in \bnf.

This comes straight from the definition of \beenf, 
but one can also see it from observing method for generating
terms in \bnf. As is shown in \cite{barendregt10}, 
this method is obtained simply by replacement of 
the first grammar rule by slightly modified rule,
resulting in following grammar.
\begin{align*}
( \pi , \Gamma )  
&\gar
(~f~( \rho_1 , \Gamma )~\dots~( \rho_m , \Gamma )~)
& \textbf{if } (f : \rho_1 \ar \dots \ar \rho_m \ar \pi) \in \Gamma
\\ 
( \sigma \rightarrow \tau , \Gamma )  
&\gar
(~\lambda~x~.~( \tau ; \Gamma,x:\sigma )~)
&   
\end{align*}

The difference lies in that $f$'s type is no longer obligated to be fully expanded
($\pi \in \T$ instead of $\alpha \in A$). This makes the grammar less deterministic,
resulting in bigger search space. The new rule is generalization of the old one,
thus all terms in \lnf will be generated, along with many new terms in \bnf that 
are not in \lnf.

By generating \lterms in \lnf we avoid generating 
\lterms $M$,$N$ such that $M \not= N$ and $M =_{\eta} N$; 
but by generating in \bnf we do not avoid this.

@@

Compare this case with corresponding grammar rule:
\[ 
	( \alpha , \Gamma )  \gar
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]

@@

Compare this case with corresponding grammar rule:
\[ 
	( \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha , \Gamma )  \gar
	\bigg( \mbox{ {\Large 
	$\lambda x_1 \dots x_n .$ 
	}}( \alpha , \Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} ) \quad \bigg)
\]
~\\
\Lets consider following $(\sigma,\Gamma)$ as a simple example:

@@

%\chapter{Ideas yet not implemented}
\chapter{Future work}

@@

%-------------------------------------------------------------------------------
%% STARY pasaže co vnich je možna eště něco co by se dalo použit
%-------------------------------------------------------------------------------
%	( \tau , \Gamma )  \gar
%	( \sigma \rightarrow \tau , \Gamma )  \gar
%	( \sigma , \Gamma )  \gar \mbox{ {\Large x}}
%	\gar & \Big( \mbox{ {\Large $\lambda$f.}}
%	\gar & 
%	\gar & 
%	\gar & 
%	\gar & 

%\section{-OLD-Inference and grammar rules producing terms in Long normal form}
%
%As is said in \cite{barendregt10} there are two
%inference/grammar rules which together generate
%simply typed lambda terms in their long normal forms.
%
%
%Inference rule 1: 
%\[
%	\frac{\Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} \vdash M : \alpha }
%	     {\Gamma \vdash (\lambda x_1 \dots x_n . M) : 
%	     \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha }
%\]
%\\
%Proof of correctness:
%\[
%	\dfrac{
%		\dfrac
%		 {\Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} \vdash M : \alpha}
%		 {\dfrac
%		   {\Gamma \cup \{ (x_1,\tau_1),\dots,(x_{n-1},\tau_{n-1})\} \cup 
%		                \{(x_n,\tau_n) \} \vdash M : \alpha}
%		   {\dfrac{\Gamma \cup \{ (x_1,\tau_1),\dots,(x_{n-1},\tau_{n-1})\}  
%		                \vdash (\lambda x_n . M) : \tau_n \rightarrow \alpha}
%				  { \vdots }		   
%		   }
%		 }		 
%	 }
%	     {\Gamma \vdash (\lambda x_1 \dots x_n . M) : 
%	     \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha }
%\]
%\\
%... there is a grammar rule:
%\[ 
%	( \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha , \Gamma )  \gar
%	\bigg( \mbox{ {\Large 
%	$\lambda x_1 \dots x_n .$ 
%	}}( \alpha , \Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} ) \quad \bigg)
%\]
%\\
%
%Inference rule 2: 
%\[
%	\frac{ (f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma \qquad
%	       \Gamma \vdash M_1 : \rho_1 \quad
%	       \dotsm \quad
%	       \Gamma \vdash M_m : \rho_m        
%	      }
%	     {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
%\]
%\\
%Proof of correctness (\textbf{TODO REPAIR} Conceptually it is ok but there is sazba-bug somewhere): 
%\[
%   \dfrac
%     {\dfrac
%      {\dfrac
%       {\dfrac         
%         {\dfrac  
%          {\dfrac
%           {\boxed{(f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma}}
%           {\Gamma \vdash f : \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha}
%           \quad
%           \boxed{\Gamma \vdash M_1 : \rho_1} }
%          {\Gamma \vdash (f M_1) : \rho_2 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha }
%          }{\vdots} 
%         \quad 
%         \ddots }
%       {\Gamma \vdash (f M_1 \dots M_{m-2}) : \rho_{m-1} \rightarrow \rho_m \rightarrow \alpha}
%       \quad
%       \boxed{\Gamma \vdash M_{m-1} : \rho_{m-1}}  }
%      {\Gamma \vdash (f M_1 \dots M_{m-1}) : \rho_m \rightarrow \alpha}       
%      \quad 
%      \boxed{\Gamma \vdash M_m : \rho_m} }
%	 {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
%\]
%\\
%... there is a grammar rule:
%\[ 
%	( \alpha , \Gamma )  \gar
%	\bigg( \mbox{ {\Large f }}
%	  \mbox{ \Vtextvisiblespace[1em] } 
%	  ( \rho_1 , \Gamma )
%	  \mbox{ \Vtextvisiblespace[1em] } 
%	  \dots
%	  \mbox{ \Vtextvisiblespace[1em] } 
%	  ( \rho_m , \Gamma )
%	  \quad \bigg)
%\]

%\begin{todo}
%& SHOW correctness of those inference rules by composing them of 
%	  $E^{\rightarrow}$, $I^{\rightarrow}$ and \textit{axiom}.
%& SHOW more examples of inference rules transformed into grammar rules.
%& DESCRIBE general algorithm for this transformation.
%& TALK ABOUT $\tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$ 
%& TALK ABOUT $\beta \eta^{-1}$-normal form which is generated by this method.
%& Napsat tu (nebo asi někam jinam
%	ale hlavně někam), že de generovat přímo i termy v beta normalní 
%	formě, ale že se mi to zdálo o dost pracnější díky tomu, že
%	musíme čekovat jestli neni něco toho typu i v ne atomickejch
%	vrcholech/typech...
%\end{todo}
